---
title: "Read In Data, Impute, Save Out"
author: "Wesley Burr"
date: "2022-11-21"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Data, Setup Packages

```{r libraries, echo = FALSE, warning = FALSE, message = FALSE, error = FALSE}
library(readr)
library(lubridate)
library(tidyverse)
library(tsinterp)
```

```{r load_clean}
dat <- readRDS("source_data.rds")

plot(dat$date, dat$`16.55`, type = "l",
     xlab = "Date/Time", ylab = "Particle Counts")
abline(v = dat$date[which(is.na(dat$`16.55`))], 
       col = "red", lty = 3)
```

### Initial Fill: Just Throw it at the Function

To start, we will just throw the missingness at the univariate
interpolator function in `tsinterp`, and let it churn. This has some
built-in logic: 

* Find all gap 'blocks'
* Linearly interpolate the gaps
* Iterate over:
    - Estimate mean trend function
    - Estimate periodic function
    - Estimate approximately stationary Wiener interpolator
      until convergence; at each stage, forward-substitute previous
      interpolation pieces, leaving only current stage unknown.

The package can be found on [https://www.github.com/wesleyburr/tsinterp](GitHub).

```{r} 
z <- dat$`16.55`
gap <- which(is.na(z))
res <- interpolate(z, gap, maxit = 20, sigClip = 0.99, delT = 60 * 60)
check <- res[[0]]
check[check < 0] <- 0.0
plot(check, type = "l")
```

This ran pretty well - there are some negatives, and the big gap at the
end wasn't very well interpolated, but it was highly erratic and variable
data, so a lot of it wasn't predictable. We could do some permutations on
the residuals and do some modeling of the permutations if there's compute 
power available.

Now, to run this for the rest of the time series in the set!

## Impute the World

```{r}
library("parallel")
```

NOTE: do not run this. It takes a very long time. The entire process on
16 cores took over 72 hours to run. Also, note that the first column
of the `dat` object is the `Date`.

```{r, eval = FALSE}
int <- mclapply(dat[, -1], FUN = function(x) { 
  z <- unlist(x) 
  gap <- which(is.na(z)) 
  res <- interpolate(z, gap, maxit = 20, sigClip = 0.99, delT = 60 * 60) 
  y <- res[[1]] 
  y[y < 0] <- 0.0 
  y 
}, mc.cores = 16)
```

We have provided the interpolated data as well.

```{r, eval = FALSE}
int <- data.frame(date = dat[, 1], int)
names(int) <- names(dat)
saveRDS(int, file = "source_data_int.rds")
```


